Collision rates of planetesimals near mean-motion resonances

### 2020-07-07 main.tex
Prepare for submission to MNRAS

\documentclass[twocolumn]{aastex63}
\usepackage{natbib}
%\definecolor{orcidlogocol}{HTML}{A6CE39}
\bibliographystyle{aasjournal}

\begin{document}

\title{COLLISION RATES OF PLANETESIMALS NEAR MEAN-MOTION RESONANCES}

\author{Spencer C. Wallace}
\affiliation{Astronomy Department, University of Washington, Seattle, WA 98195}

\author{Thomas R. Quinn}
\affiliation{Astronomy Department, University of Washington, Seattle, WA 98195}

\author{Aaron C. Boley}
\affiliation{Department of Physics and Astronomy, University of British Columbia, Vancouver BC, Canada}

\begin{abstract}
In circumstellar disks, collisional grinding of planetesimals produces second-generation dust which can be observed through thermal 
emission. While it remains unclear when second-generation dust first becomes a major component of the total dust content, the presence of 
such dust and potentially the substructure within it can be used to explore a disk's physical conditions. A perturbing planet has been shown to 
produce nonaxisymmetric structures, as well as gaps in disks, regardless of the origin of the dust. However, small grains will have very 
different dynamics compared with planetesimals when in the presence of gas, and as such, the collisional evolution of planetesimals could 
create dusty disk structures that would not exist otherwise. Previous studies have shown that morphological differences in the collisional dust 
corotating with a giant planet can be used to determine whether the planet's eccentricity is large or small. Going further, we investigate the 
production of second-generation dust near prominent mean-motion resonances with the planet using N-body simulations that directly resolve 
collisions between planetesimals. We find that a distinct over- or under-density in the dust emission is produced at the interior 2:1 mean-motion 
resonance, depending on the mass and eccentricity of the planet. The presence of one of these two features can be used to place simultaneous 
upper or lower limits on the mass and eccentricity of the perturbing body.
\end{abstract}

@@

Recent observations of circumstellar disks by ALMA have revealed a rich variety of substructure. Features such as gaps and asymmetries 
\citep{2015ApJ...808L...3A, 2016Sci...353.1519P, PhysRevLett.117.251101, 2016ApJ...820L..40A, 2016Natur.535..258C} in the emission provide 
diagnostics for the physical processes that drive the evolution of the disks. In many gas-rich systems, these gap features are argued to indicate the 
presence of a giant planet, either embedded in the disk \citep{2015MNRAS.453L..73D} or orbiting externally, although the details of gas-dust 
interactions can have strong effects on the resulting morphologies
\citep{2018ApJ...866..110D}. In gas-poor/debris systems, a giant
planet perturber can also influence 
the structure of the dust continuum emission. For example, a misaligned giant planet can produce nonaxisymmetric features such as warps 
\citep{2001A&A...370..447A}, highly eccentric perturbers can produce structures through secular interactions 
\citep{2014MNRAS.443.2541P, 2015MNRAS.448.3679P}, and mean motion resonances (MMRs) can open gaps
\citep{2015ApJ...798...83N, 2016ApJ...818..159T, 2018ApJ...857....3T}.

Collisions between planetesimals are thought to be the principal source of dust for debris disks. \citep[see][]{2008ARA&A..46..339W}.  Although some 
amount of primordial dust is likely still present during the early stages of debris disk evolution, ongoing collisions between small bodies will augment 
this and could be used to trace the underlying dynamical activity of the planetesimals. This process was directly detected when the putative object 
Fomalhaut b was confirmed to be an expanding debris cloud, most likely from a planetesimal collision \citep{2020PNAS..117.9712G}.

\citet{2013ApJ...777L..31D} showed that collisional dust that is generated near the gap opened by a giant planet in a planetesimal disk should 
produce a distinct observable marker that could be used to infer the presence of the planet. In a followup study, it was found that morphological 
differences in the dust emission could be used to determine whether the planet had a circular or eccentric orbit \citep{2016ApJ...820...29D}. Moreover, 
substructure due to mean-motion resonances (MMRs) with a giant planet may hold additional clues that could be used to constrain the properties of 
the planet.  The width of a resonance is set by both the mass of the perturbing planet and the unperturbed eccentricity of the planetesimals, 
which is in turn set by the eccentricity of the planet through secular forcing. Therefore, the structure of second generation dust produced near MMRs 
may encode information about both the mass and eccentricity of the planet.

@@

To further elucidate this subject, we use the tree based N-body code {\sc ChaNGa}\citep{2008IEEEpds...ChaNGa, 2015AphCom..2..1}  to follow the 
collisional evolution of a planetesimal disk under the gravitational influence of a Jupiter-sized body. Because particle positions are sorted into a tree 
structure, neighbor finding and collision detection can be done quickly and efficiently. This considerably relaxes the constraints on resolution and 
integration time. With this toolset, we explore the collision rate structure of a planetesimal disk in the vicinity of mean-motion resonances with the planet. In 
particular, we would like to determine (1) whether MMRs leave a detectable signature in the collisionally-generated dust, and (2) whether these 
signatures can be used to determine or constrain the orbital properties of the perturbing planet.

This work is organized in the following way: In section \ref{sec:dynamics}, we provide an overview of the relevant dynamics that drive the evolution 
of a planetesimal disk under the gravitational influence of an external perturber. In section \ref{sec:sims} we provide an overview of the N-body code 
used and describe the initial conditions chosen for five simulations in which a perturbing giant planet is given various masses 
and eccentricities. Section \ref{sec:results} presents the results of these simulations, and we use the resolved collisions to generate synthetic dust 
emission profiles that would be detected with observing facilities like ALMA. Next, we show in section \ref{sec:constrain} that a characteristic bump 
or dip feature appears in the dust emission near the interior 2:1 MMR, the presence of which depends on the mass and eccentricity of the perturbing 
planet. We discuss the potential of using this feature to place constraints on the mass and eccentricity of the planet and conclude in section 
\ref{sec:conclusions}.

@@

The most direct and widespread effect that a perturbing giant planet will have on a planetesimal disk is through secular forcing of the planetesimals. This will 
cause the complex eccentricities of the planetesimals to take on a time-independent 
forced value, given by \citep{1999ApJ...527..918W} as

@@

Without any nearby secular or mean motion-resonances, equation \ref{eq:eforced} will completely describe the eccentricities and longitude of  
pericenter orientations of the planetesimals. Additional forces due to two-body scattering between planetesimals, along with 
aerodynamic gas drag will add an additional free component to the complex eccentricity, which will be randomly oriented. The 
magnitude of the free eccentricity describes how dynamically hot the planetesimal disk is and sets the random encounter speeds of 
planetesimals. When the dynamical excitation of the disk is driven by gravitational stirring, the magnitude of the free eccentricity can 
be described by a Rayleigh distribution \citep{1992Icar...96..107I}.

@@

The net result of this is that MMRs will produce spikes in the semimajor axis-eccentricity distribution of a planetesimal disk. In order to 
preserve the Jacobi energy, equation \ref{eq:tiss} predicts that these spikes will tend to curve away from the perturbing body. This lowers the 
surface density of the disk near the locations of resonances, while producing a pileup of bodies on eccentric orbits adjacent to the MMR. The 
enhanced surface density will increase the collision rate. At the same time, the large eccentricites will either lower the collision rate due to the decreased 
effectiveness of gravitational focusing, or raise it due to the more vigorous encounter rate, depending on the magnitude of the encounter velocity relative to 
the escape velocity of the planetesimals. In addition, the MMRs cause the orbits of planetesimals to precess on relatively short 
timescales, which effectively randomizes their longitudes of perihelion. In a secularly forced disk, there will be an interface between aligned 
and randomly oriented orbits at the edge of the resonance. Due to the complicated dynamics that are at play in this region, we turn to a direct 
N-body treatment to better understand how the collision rate varies in the vicinity of a MMR.

@@

To follow the dynamical and collisional evolution of a planetesimal disk, we use the highly parallel N-body code {\sc ChaNGa} 
\footnote{A public version of {\sc ChaNGa} can be downloaded from \url{http://www-hpcc.astro.washington.edu/tools/ChaNGa.html}}. 
This code, which is written in the {\sc CHARM++} parallel programming language, was originally designed for cosmology simulations 
and has been shown to perform well on up to half a million processors \citep{2015AphCom..2..1}. {\sc ChaNGa} calculates 
gravitational forces using a modified Barnes-Hut \citep{1986Natur.324..446B} tree with hexadecapole expansions of the moments. 
All of the simulations we perform use a node opening criterion of $\theta_{BH}$ = 0.7. More information about the code can be found 
in \citet{2008IEEEpds...ChaNGa}. We have also recently modified {\sc ChaNGa} to handle solid-body collisions. A full description of 
the collision module implementation can be found in \citet{2019MNRAS.489.2159W}.

@@

In total, five simulations are run, which are listed in table
\ref{tab:sims}. The second
is a ``nominal'' case, in which the perturbing planet's mass and 
eccentricity are set to that of Jupiter's. In the other four cases, the mass (eccentricity) is increased or decreased by a factor of two, while the 
eccentricity (mass) is held at the nominal value. In all cases, the perturbing giant is placed on a 5.2 AU orbit around a 1 $M_{\odot}$ star. The 
planetesimal disk extends from 2.2 to 3.8 AU, which covers the two most prominent mean-motion resonances with the giant planet, the 2:1 at 3.27 AU 
and the 3:1 at 2.5 AU. The planetesimal disk follows a minimum-mass solar nebula surface density profile \citep{1981PThPS..70...35H}

@@

\noindent with $\alpha$ = 3/2 and $\Sigma_{0}$ = 10 g cm$^{-2}$. Planetesimals are given a bulk density of 2 g cm$^{-3}$ and a diameter of 
300 km. This corresponds to a disk containing roughly 500,000 bodies. Each simulation is evolved for 5,000 years, which is about 400 orbits 
of the perturbing planet.

@@

For the purposes of the integrator, there are two relevant timescales in this system. The first is the orbital dynamical time $\sqrt{a^3/
G M_{\odot}}$. Experimentally, we have found that using a base timestep size with {\sc ChaNGa} of 3\% of an orbital dynamical time
at the inner edge of the disk keeps the integration symplectic. Although doing so preserves orbital frequencies, the errors associated with
precession frequencies do not average out to zero. This is especially important given the effects of secular forcing by the planet. To mitigate
this, we further reduce the base timestep size by a factor of 4 and use $\Delta t$ = 0.0025 yr. Doing so prevents the longitude of perihelia of
planetesimals at the inner edge of the disk, where this artificial precession is most severe, from drifting by more than the intrinsic spread in $\varpi$ due to the
free eccentricity over the course of our integrations.

@@

We begin by examining simulations e1m2, e2m2 and e3m2. The positions of the planetesimals in the x-y plane after 5,000 years of integration are 
shown in figure \ref{fig:xy}. In all cases, the coordinate system is rotated so that the giant planet lies at $\theta = 0$. The longitude of perihelion of 
the planet is shown by the dashed line. Resonances with the perturbing planet are visible as nonaxisymmetric gaps. Upon close inspection of a 
series of simulation snapshots, these gaps appear to follow the planet in its orbit, rather than aligning themselves with the longitude of perihelion. A 
similar substructure, which is most noticeable near the 2:1 MMR, reveals itself in \citet{2000Icar..143...45R} (see bottom panel of figure 3) and 
\citet{2016ApJ...818..159T} (figure 3). This structure is also present in the \citet{2017ApJ...850..103B} simulations, although it was not reported at the 
time (see Appendix \ref{sec:boley_plot}). It is worth noting that \citet{2000Icar..143...45R} started with a completely cold planetesimal disk and a 
Jupiter mass planet on a circular, coplanar
orbit. \citet{2017ApJ...850..103B} used a similar setup.
Thus, the presence of these features seems robust 
to the choice of initial conditions.

@@

Another important effect of the resonances is visible in figure \ref{fig:long_ph}, which shows the orientation of the longitude of perihelia of 
planetesimals in the disk. Inside of the resonances, orbits of planetesimals quickly precess, and their orientations are effectively randomized. An 
important point to note is that this strong precession effect quickly disappears beyond the boundaries of the resonance. This turns out to be 
key to explaining the nonaxisymmetric structure seen in figure \ref{fig:xy}, which will be addressed in more detail below.

@@

We attribute the production of a bump or a dip at certain semi-major axes to changes in the gravitational focusing factor during two-body encounters. 
Closer to the star, the planetesimal disk starts off dynamically colder. Here, additional heating from the resonance acts to suppress the gravitational 
focusing cross section, which reduces the collision rate from its
non-perturbed value. Further out in the disk, where the unperturbed
encounter velocity is higher, the gravitational focusing cross section
is already minimized. Any additional heating instead acts to simply
increase the encounter rate between bodies, which enhances the
collision rate. This explains the relative drop in the collision rate
near the 3:1 MMR and the relative increase near the 2:1 and 5:3
resonances. Although this could potentially serve as a useful
diagnostic of the planetesimal size in a disk (because the mutual
escape velocity sets how dynamically hot the bodies can get before
gravitational focusing is suppressed), a direct measurement of the
semimajor axes of the colliding planetesimals is not
possible. Furthermore, as we will demonstrate next, collisions due to bodies in resonance do not have a significant effect on the final shape of the radial dust distribution.

\begin{figure}
\begin{center}
    \includegraphics[width=0.5\textwidth]{figures/coll_hist_a.png}
    \caption{A PDF of the collision rate in each disk as a function of semimajor axis, generated using a KDE with a Gaussian kernel with a FWHM of 0.02 AU. In semimajor axis space, prominent features appear near the 3:1 and 2:1 MMRs. Near the 3:1, the collision rate decreases from its unperturbed value, while an enhancement appears near the 2:1.\label{fig:coll_hist_a}}
\end{center}
\end{figure}

@@

Although the nonaxisymmetric pileup of bodies near the edges of resonances has been seen previously 
\citep{2000Icar..143...45R, 2016ApJ...818..159T}, we could not find a satisfactory explanation for why it happens, which we will try to provide here. As 
mentioned in section \ref{sec:dynamics}, the circulation frequency of the critical angle slows as one approaches the edge of the resonance from the 
outside. Following the pendulum analogy, this is equivalent to
approaching the point where the pendulum becomes suspended in the
at top dead center. 
When the critical angle stays relatively stationary during an encounter, the net torque will drive the longitude of pericenter towards the point of 
conjunction \citep{1976ARA&A..14..215P}.  This is the basic mechanism that causes isolated resonances to be stable. Inside of resonance, the critical 
angle librates about an equilibrium value and this configuration is maintained. Just outside of resonance, however, the critical angle slowly drifts away, with the 
drift direction changing sign across the location of exact resonance. The net result of this is that bodies just outside of the resonance experience a torque from 
the planet that temporarily drives their pericenters toward the current orbital position of the planet. After the encounter ends, differential rotation of the disk 
slowly destroys the configuration over part of an orbit. As the planet passes by, adjacent planetesimals near the resonance edges temporarily align 
themselves in this configuration. This explains why the concentration of planetesimals appears to `follow' the planet.

@@

\citet{2016ApJ...818..159T} provide a different explanation for this phenomenon, suggesting that the nonaxisymmetric structure is the product 
of the path that a resonant test particle takes in a frame co-rotating with the planet (see figure 8.4 of \citet{2000ssd..book.....M}). This explanation, 
however, does not hold for a large collection of planetesimals, especially when there is no forced eccentricity. In such a case, the orbits of 
planetesimals near resonance would be randomly aligned, and the axisymmetric structure should become washed out. When the perturbing 
planet is on a circular orbit, this structure is still present (see figure 3 of \citet{2016ApJ...818..159T} and Appendix \ref{sec:boley_plot}). The only 
connection between the structure of the gaps in the disk and the path of a resonant particle in the corotating frame is that both phenomena produce the 
same symmetries, which depend on the particular MMR.

@@

The dynamical effects of varying the mass of the planet are somewhat simpler, in that doing so does not affect the forced eccentricity of the 
planetesimal disk. Instead, only the width of the resonances changes. The width of a first order resonance scales with $m$ (see equation 
\ref{eq:res_so}), while the leading order terms in the resonant part of the disturbing function, which set the strength of the resonance, also scale as 
$m$. For the 2:1 MMR, the dynamics near the resonance are equally sensitive to changes in eccentricity and mass.

@@

While the collisional dust profiles show radial amplitude variations, we still need to assess whether those variations could be observable. To proceed, 
we create a sky model for simulated ALMA observations as follows. First, we use the radially-averaged collision profile from each simulation (see figures 
\ref{fig:coll_polar_e} and \ref{fig:coll_polar_m}) as the template for an azimuthally symmetric disk. We then scale the size of each profile by a factor of 
ten, which places the perturbing planet at 52 au. This scaling is permitted by the dynamics (both the resonance widths and the forced eccentricities 
are scale-free) and makes the disk comparable in size to the many disks that have now been observed by ALMA \citep{2018ApJ...869L..42H}. The 
angular size scale is then set by envisaging a face-on disk at a distance of 100 pc.

The sky model intensity (flux per pixel) is produced by interpolating the radial collision frequency onto a 2D Cartesian grid with cell widths of 
approximately 2 mas. In doing so, we choose to make the intensity proportional to the collision frequency (i.e., the dust) because it is the most 
straightforward case. Many different parameterizations are possible, depending on optical depth, grain size, disk temperature, etc., but we will avoid 
such complications here. We normalize the intensity by setting the total disk flux density to be 100 mJy at about 350 GHz.

>>>>

\documentclass[fleqn,usenatbib]{mnras}
\usepackage{newtxtext,newtxmath}
\usepackage[T1]{fontenc}

\DeclareRobustCommand{\VAN}[3]{#2}
\let\VANthebibliography\thebibliography
\def\thebibliography{\DeclareRobustCommand{\VAN}[3]{##3}\VANthebibliography}

\usepackage{graphicx}

\title[Collision rates of planetesimals]{Collision rates of planetesimals near mean-motion resonances}

\author[S. C. Wallace et al.]{
Spencer C. Wallace$^{1}$\thanks{E-mail: scw7@uw.edu (SCW)},
Thomas. R. Quinn$^{1}$,
Aaron C. Boley$^{2}$
\\
$^{1}$Astronomy Department, University of Washington, Seattle, WA 98195\\
$^{2}$Department of Physics and Astronomy, University of British Columbia, Vancouver BC, Canada
}

% These dates will be filled out by the publisher
\date{Accepted XXX. Received YYY; in original form ZZZ}

\pubyear{2020}

\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

\begin{abstract}
In circumstellar discs, collisional grinding of planetesimals produces second-generation dust which can be observed through thermal 
emission. While it remains unclear when second-generation dust first becomes a major component of the total dust content, the presence of 
such dust and potentially the substructure within it can be used to explore a disc's physical conditions. A perturbing planet has been shown to 
produce nonaxisymmetric structures, as well as gaps in discs, regardless of the origin of the dust. However, small grains will have very 
different dynamics compared with planetesimals when in the presence of gas, and as such, the collisional evolution of planetesimals could 
create dusty disc structures that would not exist otherwise. Previous studies have shown that morphological differences in the collisional dust 
corotating with a giant planet can be used to determine whether the planet's eccentricity is large or small. Going further, we investigate the 
production of second-generation dust near prominent mean-motion resonances with the planet using N-body simulations that directly resolve 
collisions between planetesimals. We find that a distinct over- or under-density in the dust emission is produced at the interior 2:1 mean-motion 
resonance, depending on the mass and eccentricity of the planet. The presence of one of these two features can be used to place simultaneous 
upper or lower limits on the mass and eccentricity of the perturbing body.
\end{abstract}

\begin{keywords}
planet-disc interactions -- planets and satellites: formation -- submillimetre: planetary systems -- methods: numerical
\end{keywords}

@@

Recent observations of circumstellar discs by ALMA have revealed a rich variety of substructure. Features such as gaps and asymmetries 
\citep{2015ApJ...808L...3A, 2016Sci...353.1519P, PhysRevLett.117.251101, 2016ApJ...820L..40A, 2016Natur.535..258C} in the emission provide 
diagnostics for the physical processes that drive the evolution of the discs. In many gas-rich systems, these gap features are argued to indicate the 
presence of a giant planet, either embedded in the disc \citep{2015MNRAS.453L..73D} or orbiting externally, although the details of gas-dust 
interactions can have strong effects on the resulting morphologies
\citep{2018ApJ...866..110D}. In gas-poor/debris systems, a giant
planet perturber can also influence 
the structure of the dust continuum emission. For example, a misaligned giant planet can produce nonaxisymmetric features such as warps 
\citep{2001A&A...370..447A}, highly eccentric perturbers can produce structures through secular interactions 
\citep{2014MNRAS.443.2541P, 2015MNRAS.448.3679P}, and mean motion resonances (MMRs) can open gaps
\citep{2015ApJ...798...83N, 2016ApJ...818..159T, 2018ApJ...857....3T}.

Collisions between planetesimals are thought to be the principal source of dust for debris discs. \citep[see][]{2008ARA&A..46..339W}.  Although some 
amount of primordial dust is likely still present during the early stages of debris disc evolution, ongoing collisions between small bodies will augment 
this and could be used to trace the underlying dynamical activity of the planetesimals. This process was directly detected when the putative object 
Fomalhaut b was confirmed to be an expanding debris cloud, most likely from a planetesimal collision \citep{2020PNAS..117.9712G}.

\citet{2013ApJ...777L..31D} showed that collisional dust that is generated near the gap opened by a giant planet in a planetesimal disc should 
produce a distinct observable marker that could be used to infer the presence of the planet. In a followup study, it was found that morphological 
differences in the dust emission could be used to determine whether the planet had a circular or eccentric orbit \citep{2016ApJ...820...29D}. Moreover, 
substructure due to mean-motion resonances (MMRs) with a giant planet may hold additional clues that could be used to constrain the properties of 
the planet.  The width of a resonance is set by both the mass of the perturbing planet and the unperturbed eccentricity of the planetesimals, 
which is in turn set by the eccentricity of the planet through secular forcing. Therefore, the structure of second generation dust produced near MMRs 
may encode information about both the mass and eccentricity of the planet.

@@

To further elucidate this subject, we use the tree based N-body code {\sc ChaNGa}\citep{2008IEEEpds...ChaNGa, 2015AphCom..2..1}  to follow the 
collisional evolution of a planetesimal disc under the gravitational influence of a Jupiter-sized body. Because particle positions are sorted into a tree 
structure, neighbor finding and collision detection can be done quickly and efficiently. This considerably relaxes the constraints on resolution and 
integration time. With this toolset, we explore the collision rate structure of a planetesimal disc in the vicinity of mean-motion resonances with the planet. In 
particular, we would like to determine (1) whether MMRs leave a detectable signature in the collisionally-generated dust, and (2) whether these 
signatures can be used to determine or constrain the orbital properties of the perturbing planet.

This work is organized in the following way: In section \ref{sec:dynamics}, we provide an overview of the relevant dynamics that drive the evolution 
of a planetesimal disc under the gravitational influence of an external perturber. In section \ref{sec:sims} we provide an overview of the N-body code 
used and describe the initial conditions chosen for five simulations in which a perturbing giant planet is given various masses 
and eccentricities. Section \ref{sec:results} presents the results of these simulations, and we use the resolved collisions to generate synthetic dust 
emission profiles that would be detected with observing facilities like ALMA. Next, we show in section \ref{sec:constrain} that a characteristic bump 
or dip feature appears in the dust emission near the interior 2:1 MMR, the presence of which depends on the mass and eccentricity of the perturbing 
planet. We discuss the potential of using this feature to place constraints on the mass and eccentricity of the planet and conclude in section 
\ref{sec:conclusions}.

@@

The most direct and widespread effect that a perturbing giant planet will have on a planetesimal disc is through secular forcing of the planetesimals. This will 
cause the complex eccentricities of the planetesimals to take on a time-independent 
forced value, given by \citep{1999ApJ...527..918W} as

@@

Without any nearby secular or mean motion-resonances, equation \ref{eq:eforced} will completely describe the eccentricities and longitude of  
pericenter orientations of the planetesimals. Additional forces due to two-body scattering between planetesimals, along with 
aerodynamic gas drag will add an additional free component to the complex eccentricity, which will be randomly oriented. The 
magnitude of the free eccentricity describes how dynamically hot the planetesimal disc is and sets the random encounter speeds of 
planetesimals. When the dynamical excitation of the disc is driven by gravitational stirring, the magnitude of the free eccentricity can 
be described by a Rayleigh distribution \citep{1992Icar...96..107I}.

@@

The net result of this is that MMRs will produce spikes in the semimajor axis-eccentricity distribution of a planetesimal disc. In order to 
preserve the Jacobi energy, equation \ref{eq:tiss} predicts that these spikes will tend to curve away from the perturbing body. This lowers the 
surface density of the disc near the locations of resonances, while producing a pileup of bodies on eccentric orbits adjacent to the MMR. The 
enhanced surface density will increase the collision rate. At the same time, the large eccentricites will either lower the collision rate due to the decreased 
effectiveness of gravitational focusing, or raise it due to the more vigorous encounter rate, depending on the magnitude of the encounter velocity relative to 
the escape velocity of the planetesimals. In addition, the MMRs cause the orbits of planetesimals to precess on relatively short 
timescales, which effectively randomizes their longitudes of perihelion. In a secularly forced disc, there will be an interface between aligned 
and randomly oriented orbits at the edge of the resonance. Due to the complicated dynamics that are at play in this region, we turn to a direct 
N-body treatment to better understand how the collision rate varies in the vicinity of a MMR.

@@

To follow the dynamical and collisional evolution of a planetesimal disc, we use the highly parallel N-body code {\sc ChaNGa} 
\footnote{A public version of {\sc ChaNGa} can be downloaded from \url{http://www-hpcc.astro.washington.edu/tools/ChaNGa.html}}. 
This code, which is written in the {\sc CHARM++} parallel programming language, was originally designed for cosmology simulations 
and has been shown to perform well on up to half a million processors \citep{2015AphCom..2..1}. {\sc ChaNGa} calculates 
gravitational forces using a modified Barnes-Hut \citep{1986Natur.324..446B} tree with hexadecapole expansions of the moments. 
All of the simulations we perform use a node opening criterion of $\theta_{BH}$ = 0.7. More information about the code can be found 
in \citet{2008IEEEpds...ChaNGa}. We have also recently modified {\sc ChaNGa} to handle solid-body collisions. A full description of 
the collision module implementation can be found in \citet{2019MNRAS.489.2159W}.

@@

In total, five simulations are run, which are listed in table
\ref{tab:sims}. The second
is a ``nominal'' case, in which the perturbing planet's mass and 
eccentricity are set to that of Jupiter's. In the other four cases, the mass (eccentricity) is increased or decreased by a factor of two, while the 
eccentricity (mass) is held at the nominal value. In all cases, the perturbing giant is placed on a 5.2 AU orbit around a 1 $M_{\odot}$ star. The 
planetesimal disc extends from 2.2 to 3.8 AU, which covers the two most prominent mean-motion resonances with the giant planet, the 2:1 at 3.27 AU 
and the 3:1 at 2.5 AU. The planetesimal disc follows a minimum-mass solar nebula surface density profile \citep{1981PThPS..70...35H}

@@

\noindent with $\alpha$ = 3/2 and $\Sigma_{0}$ = 10 g cm$^{-2}$. Planetesimals are given a bulk density of 2 g cm$^{-3}$ and a diameter of 
300 km. This corresponds to a disc containing roughly 500,000 bodies. Each simulation is evolved for 5,000 years, which is about 400 orbits 
of the perturbing planet.

@@

For the purposes of the integrator, there are two relevant timescales in this system. The first is the orbital dynamical time $\sqrt{a^3/
G M_{\odot}}$. Experimentally, we have found that using a base timestep size with {\sc ChaNGa} of 3\% of an orbital dynamical time
at the inner edge of the disc keeps the integration symplectic. Although doing so preserves orbital frequencies, the errors associated with
precession frequencies do not average out to zero. This is especially important given the effects of secular forcing by the planet. To mitigate
this, we further reduce the base timestep size by a factor of 4 and use $\Delta t$ = 0.0025 yr. Doing so prevents the longitude of perihelia of
planetesimals at the inner edge of the disc, where this artificial precession is most severe, from drifting by more than the intrinsic spread in $\varpi$ due to the
free eccentricity over the course of our integrations.

@@

We begin by examining simulations e1m2, e2m2 and e3m2. The positions of the planetesimals in the x-y plane after 5,000 years of integration are 
shown in figure \ref{fig:xy}. In all cases, the coordinate system is rotated so that the giant planet lies at $\theta = 0$. The longitude of perihelion of 
the planet is shown by the dashed line. Resonances with the perturbing planet are visible as nonaxisymmetric gaps. Upon close inspection of a 
series of simulation snapshots, these gaps appear to follow the planet in its orbit, rather than aligning themselves with the longitude of perihelion. A 
similar substructure, which is most noticeable near the 2:1 MMR, reveals itself in \citet{2000Icar..143...45R} (see bottom panel of figure 3) and 
\citet{2016ApJ...818..159T} (figure 3). This structure is also present in the \citet{2017ApJ...850..103B} simulations, although it was not reported at the 
time (see Appendix \ref{sec:boley_plot}). It is worth noting that \citet{2000Icar..143...45R} started with a completely cold planetesimal disc and a 
Jupiter mass planet on a circular, coplanar
orbit. \citet{2017ApJ...850..103B} used a similar setup.
Thus, the presence of these features seems robust 
to the choice of initial conditions.

@@

Another important effect of the resonances is visible in figure \ref{fig:long_ph}, which shows the orientation of the longitude of perihelia of 
planetesimals in the disc. Inside of the resonances, orbits of planetesimals quickly precess, and their orientations are effectively randomized. An 
important point to note is that this strong precession effect quickly disappears beyond the boundaries of the resonance. This turns out to be 
key to explaining the nonaxisymmetric structure seen in figure \ref{fig:xy}, which will be addressed in more detail below.

@@

We attribute the production of a bump or a dip at certain semi-major axes to changes in the gravitational focusing factor during two-body encounters. 
Closer to the star, the planetesimal disc starts off dynamically colder. Here, additional heating from the resonance acts to suppress the gravitational 
focusing cross section, which reduces the collision rate from its
non-perturbed value. Further out in the disc, where the unperturbed
encounter velocity is higher, the gravitational focusing cross section
is already minimized. Any additional heating instead acts to simply
increase the encounter rate between bodies, which enhances the
collision rate. This explains the relative drop in the collision rate
near the 3:1 MMR and the relative increase near the 2:1 and 5:3
resonances. Although this could potentially serve as a useful
diagnostic of the planetesimal size in a disc (because the mutual
escape velocity sets how dynamically hot the bodies can get before
gravitational focusing is suppressed), a direct measurement of the
semimajor axes of the colliding planetesimals is not
possible. Furthermore, as we will demonstrate next, collisions due to bodies in resonance do not have a significant effect on the final shape of the radial dust distribution.

\begin{figure}
\begin{center}
    \includegraphics[width=0.5\textwidth]{figures/coll_hist_a.png}
    \caption{A PDF of the collision rate in each disc as a function of semimajor axis, generated using a KDE with a Gaussian kernel with a FWHM of 0.02 AU. In semimajor axis space, prominent features appear near the 3:1 and 2:1 MMRs. Near the 3:1, the collision rate decreases from its unperturbed value, while an enhancement appears near the 2:1.\label{fig:coll_hist_a}}
\end{center}
\end{figure}

@@

Although the nonaxisymmetric pileup of bodies near the edges of resonances has been seen previously 
\citep{2000Icar..143...45R, 2016ApJ...818..159T}, we could not find a satisfactory explanation for why it happens, which we will try to provide here. As 
mentioned in section \ref{sec:dynamics}, the circulation frequency of the critical angle slows as one approaches the edge of the resonance from the 
outside. Following the pendulum analogy, this is equivalent to
approaching the point where the pendulum becomes suspended in the
at top dead center. 
When the critical angle stays relatively stationary during an encounter, the net torque will drive the longitude of pericenter towards the point of 
conjunction \citep{1976ARA&A..14..215P}.  This is the basic mechanism that causes isolated resonances to be stable. Inside of resonance, the critical 
angle librates about an equilibrium value and this configuration is maintained. Just outside of resonance, however, the critical angle slowly drifts away, with the 
drift direction changing sign across the location of exact resonance. The net result of this is that bodies just outside of the resonance experience a torque from 
the planet that temporarily drives their pericenters toward the current orbital position of the planet. After the encounter ends, differential rotation of the disc 
slowly destroys the configuration over part of an orbit. As the planet passes by, adjacent planetesimals near the resonance edges temporarily align 
themselves in this configuration. This explains why the concentration of planetesimals appears to `follow' the planet.

@@

\citet{2016ApJ...818..159T} provide a different explanation for this phenomenon, suggesting that the nonaxisymmetric structure is the product 
of the path that a resonant test particle takes in a frame co-rotating with the planet (see figure 8.4 of \citet{2000ssd..book.....M}). This explanation, 
however, does not hold for a large collection of planetesimals, especially when there is no forced eccentricity. In such a case, the orbits of 
planetesimals near resonance would be randomly aligned, and the axisymmetric structure should become washed out. When the perturbing 
planet is on a circular orbit, this structure is still present (see figure 3 of \citet{2016ApJ...818..159T} and Appendix \ref{sec:boley_plot}). The only 
connection between the structure of the gaps in the disc and the path of a resonant particle in the corotating frame is that both phenomena produce the 
same symmetries, which depend on the particular MMR.

@@

The dynamical effects of varying the mass of the planet are somewhat simpler, in that doing so does not affect the forced eccentricity of the 
planetesimal disc. Instead, only the width of the resonances changes. The width of a first order resonance scales with $m$ (see equation 
\ref{eq:res_so}), while the leading order terms in the resonant part of the disturbing function, which set the strength of the resonance, also scale as 
$m$. For the 2:1 MMR, the dynamics near the resonance are equally sensitive to changes in eccentricity and mass.

@@

While the collisional dust profiles show radial amplitude variations, we still need to assess whether those variations could be observable. To proceed, 
we create a sky model for simulated ALMA observations as follows. First, we use the radially-averaged collision profile from each simulation (see figures 
\ref{fig:coll_polar_e} and \ref{fig:coll_polar_m}) as the template for an azimuthally symmetric disc. We then scale the size of each profile by a factor of 
ten, which places the perturbing planet at 52 au. This scaling is permitted by the dynamics (both the resonance widths and the forced eccentricities 
are scale-free) and makes the disc comparable in size to the many discs that have now been observed by ALMA \citep{2018ApJ...869L..42H}. The 
angular size scale is then set by envisaging a face-on disc at a distance of 100 pc.

The sky model intensity (flux per pixel) is produced by interpolating the radial collision frequency onto a 2D Cartesian grid with cell widths of 
approximately 2 mas. In doing so, we choose to make the intensity proportional to the collision frequency (i.e., the dust) because it is the most 
straightforward case. Many different parameterizations are possible, depending on optical depth, grain size, disc temperature, etc., but we will avoid 
such complications here. We normalize the intensity by setting the total disc flux density to be 100 mJy at about 350 GHz.

The simulated observations are performed using the {\sc simobserve} task in {\sc CASA} \citep{2007ASPC..376..127M}.  The disc is given an 
$\rm RA = 11^{\rm h} 01^{\rm m} 52^{\rm s}$ and a $\delta = -34^\circ 42\arcmin 17\arcsec$ (i.e., the J2000 coordinates for TW Hydrae) and is 
``observed'' through transit on March 20, 2020. We observe the disc using configurations 8, 9, and 10 at 350 GHz (cycle 6 antenna file), spending six 
hours on-source in each configuration\footnote{We have imagined that the TAC really likes us.}.  The visibilities are corrupted with thermal noise, 
setting the precipitable water vapor to 1.5 mm. They are then combined, imaged, and cleaned
using the task {\tt tclean}. Imaging uses Briggs weighting with a robustness 
parameter of -1.

The results for the nominal simulation (e2m2) are shown in figure \ref{fig:alma_sim_obs}. The 3:1 and 2:1 resonances can be easily identified in the 
cleaned image and in the corresponding radial profile.  Moreover, the features are qualitatively similar to the bright rings and dark gaps seen in actual 
disc profiles \citep{2015ApJ...808L...3A}.
